Pangenome graph layout by Path-Guided Stochastic Gradient Descent

### 2024-02-08 manuscript/main.tex
gimme more

	A graph layout is the arrangement of nodes and edges in an \textit{N}-dimensional space.
	Graph layout algorithms aim to find optimal node coordinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote an intuitive understanding of the graph.
	One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which produces aesthetic layouts.
	This is prone to get stuck in local minima, but stochastic gradient descent (SGD) implementations alleviate such a problem \citep{Zheng2019}.
	SGD uses the gradient of its individual terms to approximate the gradient of a sum of functions.

@@

	Here, we present a new pangenome graph layout algorithm which applies a path-guided stochastic gradient descent (PG-SGD) to use the paths as an embedded positional system to find distances between nodes, moving pairs of nodes in parallel with a modified HOGWILD! strategy.
	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.
	To our knowledge, no algorithm takes into account such biological information to compute the graph layouts.
	PG-SGD can be extended in any number of dimensions.
	In the ODGI toolkit \citep{Guarracino2022}, we provide implementations for 1-dimensional (1D) and 2-dimensional (2D) layouts.
	These algorithms have already been successfully applied to construct and visualize large-scale pangenome graphs of the Human Pangenome Reference Consortium (HPRC)~\citep{Liao2023, Guarracino2023}.

@@

	We presented Path-Guided Stochastic Gradient Descent (PG-SGD), the first layout algorithm for pangenome graphs that leverages the biological information available within the genomes represented in the graph.
	Our implementation efficiently computes the layout of pangenome graphs representing thousands of whole genomes.

>>>>

	A graph layout is the arrangement of nodes and edges in an \textit{N}-dimensional space.
	Graph layout algorithms aim to find optimal node coordinates in order to minimize overlapping nodes or edges, reduce edge crossings, and promote an intuitive understanding of the graph.
	One popular approach is force-directed graph drawing \citep{cheong_force-directed_2022} which produces aesthetic layouts. %by ....
	This is prone to get stuck in local minima, but stochastic gradient descent (SGD) implementations alleviate such a problem \citep{Zheng2019}.
	SGD uses the gradient of its individual terms to approximate the gradient of a sum of functions.

@@

	Here, we present a new pangenome graph layout algorithm which applies a path-guided stochastic gradient descent (PG-SGD) to use the paths as an embedded positional system to find distances between nodes, moving pairs of nodes in parallel with a modified HOGWILD! strategy.
	The algorithm computes the pangenome graph layout that best reflects the nucleotide sequences in the graph.
	\red{To our knowledge, no generic graph layout algorithm takes into account such path encoded biological information when computing a graph's layout.}
	PG-SGD can be extended in any number of dimensions.
	In the ODGI toolkit \citep{Guarracino2022}, we provide implementations for 1-dimensional (1D) and 2-dimensional (2D) layouts.
	These algorithms have already been successfully applied to construct and visualize large-scale pangenome graphs of the Human Pangenome Reference Consortium (HPRC)~\citep{Liao2023, Guarracino2023}.

@@

	\red{We presented Path-Guided Stochastic Gradient Descent (PG-SGD), the first layout algorithm for pangenome graphs that leverages the biological information available within the genomes represented in the graph. 
	Other generic graph layout algorithms, such as the one offered by \textit{BandageNG}, ignore this additional information.}
	Our implementation efficiently computes the layout of pangenome graphs representing thousands of whole genomes.
